---
title: "Using GeoProfile"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{geoprofile}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

- What is geoprofiling?
- Where does it come from?
- Where else can it be applied?
- Why should you use it?

## Loading data

First we must load the data and format it correctly for use in the `geoParams()` function. Here we will use the example data of crime in North London.

In the case of the example data these are actually already in the correct format, however the relevant code is included for completeness.
```{r}
library(GeoProfile)
d <- LondonExample_crimes
s <- LondonExample_sources

d <- geoData(d$longitude, d$latitude)
s <- geoDataSource(s$longitude, s$latitude)
```

### Explaining the data

This data comes in two parts:

1. `d` a set of coordinates representing events that we have observed.
2. `s` a set of coordinates representing sources of these events.

Of these two dataframes, only `d` is used for the model fitting. In most cases this is the only data you are working with.

The data in `s` could be a suspected set of places that are the source of the activity.

Given the origin of geoprofiling as a technique (in criminology), the example data here is a set of crime reports `d` and a set of potential "hideouts" `s`.

If the posterior distribution overlaps with a point in `s` then we can say it is quite likely that the perpetrators of said crime are located in that hideout.

## Model parameterisation

Once the data is loaded, we must set up the model and MCMC (Monte Carlo Markov Chain) parameters (for full details see the `geoParams()` help file).

For this run let's set up with some sensible sigma settings (sigma represents the standard deviation of the dispersal distribution in km).

You can parameterise the prior of sigma in one of three ways:

- Defining `sigma_mean` and `sigma_var`
- Defining `sigma_mean` and `sigma_squared_shape`
- Defining `sigma_squared_shape` only

As an example, if using `sigma_mean` and `sigma_squared_shape` to parameterise, the exact value of `sigma_var` derived here is calculated as follows:


$$
\alpha = \: \text{sigma_squared_shape}
$$

$$
\beta = \exp{{\left(2\ln{\bar{\sigma}} + 2\ln{|\,\Gamma(\alpha)|} + 2\ln{|\,\Gamma(\alpha - 0.5)|}\right)}}
$$

$$
\epsilon = \sqrt{\beta} \:\frac{\Gamma(\alpha - 0.5)}{\Gamma(\alpha)}
$$
$$
\text{sigma_var} = \frac{\beta}{(\alpha - 1) - \epsilon^2}
$$

For the MCMC we'll use 5 chains, discarding 1,000 iterations of burnin and sampling 10,000 iterations to generate the final posterior distribution.

``` {r}
# set model and MCMC parameters
p = gp.params(
  data = d,
  sigma_mean = 1,
  sigma_squared_shape = 2,
  chains = 5,
  burnin = 1e3, 
  samples = 1e4
  )
```

And let's just check to see what our full parameters object looks like:

```{r}
p
```


## Running the MCMC model

Now we are ready to actually fit the geoprofile model using an MCMC approach.

Posterior draws are smoothed to produce a posterior surface, and converted into a geoProfile.

The outputs of this function include posterior draws of alpha and sigma under the variable-sigma model.
``` r
m = geoMCMC(data = d, params = p)
```

```{r echo=FALSE}
m = geoMCMC(data = d, params = p, smoothprogress = FALSE)
```

_Note: we can control the bandwidth of the posterior smoothing by specifying the `lambda` argument. If this is not specified, the optimal bandwidth is chosen via maximum-likelihood._

Let's plot the prior and posterior of sigma. Remember that sigma is the standard deviation of the dispersal distribution, so a larger posterior value is more varied dispersal in the data.
``` {r}
geoPlotSigma(params = p, mcmc = m)
```

## Plotting

Using the wonderful `leaflet` package, we can plot the posterior surfaces on a map!

Here the red dots are crime reports, the blue dots are potential sources, and the heatmap is the posterior probability surface.

``` {r}
# plot profile on map
mapGP <- geoPlotLeaflet(params = p, data = d, source = s, surface = m$geoProfile)
mapGP
```

We can also retrieve hitscores for our potential sources (measures of how likely a potential source is to be the actual source of observed events).

``` {r}
# get hitscores
hs <- geoReportHitscores(params = p, source = s, surface =m$geoProfile)
hs
```
And here is a Lorenz plot for the same data.

``` {r}
# produce Lorenz plot
Gini <- geoPlotLorenz(hit_scores = hs, crimeNumbers = NULL)
Gini
```

``` {r}
# zoom
# FW: Currently does not work due to incorrect arguments

# zoomLon = c(-0.1, -0.01)
# zoomLat = c(51.51, 51.54)
# mapZoom <- geoPlotLeaflet(lonLimits = zoomLon, latLimits = zoomLat, params = p,
#                 data = d, source = s, surface = m$geoProfile)
# mapZoom
```

We can also plot posterior allocation and co-allocation

``` {r}
# plot allocation
geoPlotAllocation(mcmc = m)
```

``` {r}
# plot co-allocation
geoPlotCoallocation(mcmc = m)
```

3D plots are almost entirely unhelpful from a data-communication standpoint (unless interactive and animated), however they can look nice in decorative roles.

You can create these using `geoPersp()`function.

``` {r}
# produce perspective plots
# probabilities
geoPersp(surface = m$posteriorSurface, aggregate_size = 3, surface_type = "prob")
```

``` {r}
# ranked surface
geoPersp(surface = m$geoProfile, aggregate_size = 3)
```

A better way of visualising data in 3D is to use an interactive plot using the `geoSurface3D()` function.

This gets around the problem of data occlusion, though is still mostly useful in a decorative scenario.

_Note: `geoSurface3D()` requires the `plotly` package to be installed._

This particular representation has been downsampled for this example to save on resources.

```{r include=FALSE}
downsample_mat <- function(m, window = 5) {
  # Clunky function, assumes a square matrix
  if (nrow(m) %% window != 0 || ncol(m) %% window != 0) {
    stop("Matrix is not square or window is not a full divisor of the matrix size!")
  }
  
  outmat = matrix(NA, ncol = ncol(m) / window, nrow = nrow(m) / window)
  # Generate list of sampling points
  startpoints <- seq(1, nrow(m), window)
  endpoints <- startpoints + (window - 1)
  for (i in 1:length(startpoints)) {
    for (j in 1:length(startpoints)) {
      outmat[i, j] <- mean(m[startpoints[i]:endpoints[i], startpoints[j]:endpoints[j]])
    }
  }
  outmat / max(outmat)
}
downsampled <- downsample_mat(m$geoProfile, 25)
```

```r
geoSurface3D(m$geoProfile, "gp")
```
```{r echo=FALSE}
geoSurface3D(downsampled, "gp")
```


We can also get the centroids of the data split by the posterior groups. This is a coarse indication of best-guess source locations.

``` {r}
# find centroids of data split by best grouping (placeholder for more thorough method)
ms <- geoModelSources(mcmc = m, data = d)
ms
```

``` {r}
# add peaks to map
# NB requires ggplot2
# FW: mapGP is no longer a ggplot object (as it is now a leaflet object!) This needs a fix.
# library(ggplot2)
# mapSource <- mapGP + geom_point(aes(ms$longitude,ms$latitude), size=6, pch = 3, col="red")
# mapSource
```

Finally (just for fun) we can plot the posterior surface in the style of the famous Unknown Pleasures album cover by Joy Division.

``` {r}
# plot surface in style of 'unknown pleasures', for fun
unknownPleasures(m$geoProfile, paper_ref = "RgeoProfile 2.1.0")
```

## Benchmarking

Alternative methods for solving this problem include the ring search approach (searching in circles of iteratively expanding radii around each observation).

This approach is implemented in the `geoRing()` function, which can be plotted in the same manner as `geoMCMC()` output.

``` {r}
#------------------------------------------------------------------
# compare to alternative ring search strategy
#------------------------------------------------------------------
# compare to geoprofile based on ring search strategy
surface_ring <- geoRing(params = p, data = d, source = s, mcmc = m)
gp_ring <- geoProfile(surface = surface_ring)
# map of ring search geoprofile
mapRing <- geoPlotLeaflet(params = p, data = d, source = s, surface = gp_ring,
                surfaceCols = c("red", "white"))
mapRing
```

``` {r}
# hitscores of ring search geoprofile
hs_ring <- geoReportHitscores(params = p, source = s, surface = gp_ring)
hs_ring
```

## Incorporating GIS data

We may wish to use GIS data to modify the geoprofile in some manner. For example to exclude areas not on land, or weight the probabilities within a specific postcode differently.

We can do this using the `geoMask()` function.

``` {r}
#------------------------------------------------------------------
# incorporate GIS data
#------------------------------------------------------------------
# read in north London shapefile as mask and adjust surface
north_london_mask <- geoShapefile()

# restrict mask to Tower Hamlets
TH_mask <- north_london_mask[which(north_london_mask$NAME == "Tower Hamlets"),]
prob_masked <- geoMask(probSurface = m$posteriorSurface, params = p, mask = TH_mask,
                operation = "outside", scaleValue = 1e-9)
gp_masked <- geoProfile(prob_masked$prob)
```

``` {r}
# plot new surface
mapMask <- geoPlotLeaflet(params = p, data = d, source = s, surface = gp_masked)

# FW: this part does not currently work
# breakPercent = seq(0,25,l = 11))
mapMask

# hs of masked surface
hs_mask <- geoReportHitscores(params = p, source = s, surface = gp_masked)
hs_mask
```


